---
editor_options:
  chunk_output_type: console
output:
  pdf_document: default
  html_document: default
---
# Libraries: What Drives Visits And What Visitors Do by Marie Jordan

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(plyr)
library(dplyr)
library(tidyr)
library(gridExtra)
library(GGally)



```

```{r echo=FALSE, message=TRUE, warning=TRUE}
# Load the Data
master <- read.csv('PLS_FY2015_State_pusum15a.csv', check.names = F)
```

The dataset I chose is the Public Library State Summary/State Characteristics
data file from FY2015, provided by the Institute of Museum and Library Services.
The information comes from all public libraries in the United States
including Washington DC, Guam and American Samoa. The data was accompanied by 
an extensive document detailing collection techniques and data manipulation, as
well as defining the terms used in the table.

It is worth noting that this dataset is collected data from the entire 
population of public libraries in the US. It includes imputations for non-
responding libraries to minimize the effects of nonresponse. Other relevant data
is that there are different reporting periods from state to state (although all 
provided data for at least 12 months) and that there is different adherence to 
the definitions of collected data from state to state.

# Univariate Plots Section

```{r create physical df, echo=FALSE, message=FALSE, warning=FALSE}
#create physical df
physical <- subset(master, select = c('STABR', 'POPU_ST', 'VISITS', 'TOTPRO', 
                                     'HRS_OPEN'))
#remove scientific notation
options(scipen=999)
```


```{r Histograms of visit data, echo=FALSE, message=FALSE, warning=FALSE}

a1 <- ggplot(aes(x=POPU_ST), data=physical) + 
  geom_histogram() +
  labs(x='State Population', y='Count')

a2 <- ggplot(aes(x=VISITS), data=physical) + 
  geom_histogram() +
  labs(x='Number of Visits', y='Count')

a3 <- ggplot(aes(x=TOTPRO), data=physical) + 
  geom_histogram() +
  labs(x='Number of Programs', y='Count')

a4 <- ggplot(aes(x=HRS_OPEN), data=physical) + 
  geom_histogram() +
  labs(x='Number of Hours Open', y='Count')

grid.arrange(a1, a2, a3, a4, ncol=1)
```

This is a quick-and-dirty look for any interesting patterns in the variables
I'm examining that are related to visits to the libraries.

```{r Plots of visit data minus five percent outliers, echo=FALSE, message=FALSE, warning=FALSE}

summary(physical)

```

Summary table for visit-related variables. 

```{r Plots of visit data minus ten percent outliers, echo=FALSE, message=FALSE, warning=FALSE}

b1 <- ggplot(aes(x=POPU_ST), 
  data=subset(physical, physical$POPU_ST < quantile(physical$POPU_ST, 0.95) &   physical$POPU_ST > quantile(physical$POPU_ST, .05))) + 
  geom_histogram() +
  labs(x='State Population', y='Count')

c1 <- ggplot(aes(x=POPU_ST), data=subset(physical, physical$POPU_ST < quantile(physical$POPU_ST, 0.90) & physical$POPU_ST > quantile(physical$POPU_ST, .1))) + 
  geom_histogram() +
  labs(x='State Population', y='Count')

grid.arrange(a1, b1, c1, ncol=1)
```

These plots show state population numbers with outliers removed. (In this case,
 because the data is arranged about the same across all variables, population is
  acting as a test case.) The topmost plot has all 53 datapoints, the middle 
  plot removes the top and bottom 5%, and the bottom plot removes the top and
   bottom 10%.

```{r Plots of Visitor Actions, echo=FALSE, message=FALSE, warning=FALSE}
action <- subset(master, 
select = c('STABR', 'POPU_ST', 'TOTCIR', 'KIDCIRCL', 'ELMATCIR', 'REFERENC', 
'LOANTO', 'LOANFM'))

# This creates my dataframe of variables related to (some of) what visitors
#do once they're in the library.

d1 <- ggplot(aes(x=TOTCIR), data=action) + 
  geom_histogram() +
  labs(x='Total Circulation', y='Count')

d2 <- ggplot(aes(x=KIDCIRCL), data=action) + 
  geom_histogram() +
  labs(x='Children\'s Circulation', y='Count')

d3 <- ggplot(aes(x=ELMATCIR), data=action) + 
  geom_histogram() +
  labs(x='Electronic Circulation', y='Count')

d4 <- ggplot(aes(x=REFERENC), data=action) + 
  geom_histogram() +
  labs(x='Reference Circulation', y='Count')

d5 <- ggplot(aes(x=LOANTO), data=action) + 
  geom_histogram() +
  labs(x='Books Loaned Out', y='Count')

d6 <- ggplot(aes(x=LOANFM), data=action) + 
  geom_histogram() +
  labs(x='Books Loaned In', y='Count')

grid.arrange(d1, d2, d3, d4, d5, d6, ncol=1)


```

To cover visitor interactions at the library, I created a dataframe with state 
names, population numbers, and variables related to visitor actions in the 
libraries.

These tables are a quick-and-dirty look for interesting patterns. As expected, circulation of electronic materials and kid's materials is less than the total 
number. Loans to and from seem to echo each other in very interesting ways -- 
states seem to have the same numbers for loans to and loans from, for the most 
part. 

```{r summary of action, echo=FALSE, message=FALSE, warning=FALSE }

summary(action)

```

Summary of the table of actions patrons took at the library.

```{r Plot of Total Circulation minus top and bottom ten percent, echo=FALSE, message=FALSE, warning=FALSE}


e1 <- ggplot(aes(x=TOTCIR), data=subset(action, action$TOTCIR < quantile(action$TOTCIR, 0.90) & 
action$TOTCIR > quantile(action$TOTCIR, .1))) + 
  geom_histogram() +
  labs(x='Total Circulation', y='Count')

grid.arrange(d1, e1, ncol=1)

```

As with the physical data subset, there are huge differences between the min
and max, and the 3rd quartile and max. The plot shows the dataset for total 
circulation with the top and bottom 10% of states removed, to check for patterns
 without outliers.

# Univariate Analysis

### What is the structure of your dataset?

My dataset is large, with 53 observations of 124 different variables. It's 
arranged by state, with each row containing the state name, and then all of
the related information.

For this analysis, I'm concentrating on number of visits, and what visitors do
once they're in the libraries. The dataset is notable for the huge range of 
data it contains -- I could as easily have looked at funding streams, wages
and benefits, expenses, population served, or systems of main and branch 
libraries and the staffing patterns found therein.

### What is/are the main feature(s) of interest in your dataset?

My data explorations are going to tackle two primary questions -- what drives 
number of visits (focusing on number of programs and hours open in a 
year), and what they do once they're there (borrow books or electronic resources
, use the library for their children or themselves, file reference requests, 
etc.) I am examining these numbers broken down by state.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The variables I've chosen to look at directly relate to the two major interest 
features I'm examining. Although there are likely other features in the dataset
 that have influence, I've chosen to primarily look at hours open and programs 
 put on to influence visit numbers, and circulation, reference requests and 
 loans as aspects of what library users actually do.
 
### Did you create any new variables from existing variables in the dataset?

I did not create any new variables at this stage, nor did I transform or tidy
the data, beyond creating two new dataframes. For my visit numbers information, 
I created a new dataframe called physical which breaks out the variables I'm 
interested in. This makes working with the data easier, especially for things 
like summaries. I performed a similar action for visitor activities, creating a 
new dataframe called action, for the same reasons I created the physical 
dataframe. 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

With my analyses, I did not find any
particularly unusual distributions, only that there was a huge difference 
between minimum and maximum numbers for all variables studied, as would be 
expected when working with state-level data. The majority of the data for all
variables was distributed roughly evenly clustered at the lower end of the
scale, with only a few larger outliers, generally in high-population states.

Removing the outliers did not reveal any new patterns or throw up anything of 
particular interest.

# Bivariate Plots Section

```{r Correlation Plots, echo=FALSE, message=FALSE, warning=FALSE}
ggcorr(physical, nbreaks=5)

```

```{r Correlation Plots 2, echo=FALSE, message=FALSE, warning=FALSE}
ggcorr(action, nbreaks = 5)
```


Correlation plots show potential correlations between all variables 
examined.

```{r Plots of populations against visits and programs, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=POPU_ST, y= VISITS), data=physical) +
  geom_point() +
  labs(x="State Population", y="No. of Visits")

```

```{r Plots of populations against visits and programs 2, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=POPU_ST, y=TOTPRO), data=physical) +
  geom_point() +
  labs(x="State Population", y="No. of Programs")
```

These plots show the relationship between state populations and the number of 
visits and number of programs in the submitted timeframe.

```{r Number of Programs vs Number of Visits, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=TOTPRO, y = VISITS), data=physical) +
  geom_point() +
  labs(x="No. of Visits", y="No. of Programs")


```

```{r Number of Programs vs Number of Visits 2, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=log10(TOTPRO), y = log10(VISITS)), data=physical) +
  geom_point() +
  labs(x="No. of Visits", y="No. of Programs")

```


These plots examine the relationship between the number of programs and the
number of visits in a state. I did a version with the log10 of both variables,
despite the obvious linear regression in the first graph, simply because I'm
still learning how log10 works and I was curious to see what effect, if any, it
would have on the data.

```{r Number of Hours Open vs Number of Visits, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=HRS_OPEN, y = VISITS), data=physical) +
  geom_point() +
  labs(x="No. of Hours Open", y="No. of Visits")

```

As above, this plot examines the relationship between hours open and number of
visits.

```{r Linear Regressions for programs and hours open, echo=FALSE, message=FALSE, warning=FALSE}

lin_program <- lm(VISITS ~ TOTPRO, data=physical)

summary(lin_program)

lin_hours <- lm(VISITS ~ HRS_OPEN, data=physical)

summary(lin_hours)
```

Linear regression modeling confirms a significant relationship between both 
number of programs and hours open and number of visits.

```{r total predicts kid circ, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=TOTCIR, y=KIDCIRCL), data = action) +
  geom_point(size=5) +
  labs(x="Total Circulation", y="Children\'s Circulation")

```

This plot compares how total circulation numbers predict kid circulation 
numbers. There's a clear linear relationship between the two, with kid 
circulation going up as total circulation does.

```{r total predicts electronic circ, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=TOTCIR, y=ELMATCIR), data = action) +
  geom_point(size=5) +
  labs(x="Total Circulation", y="Electronic Circulation")

```

Similar to the plot above, this one looks at how total circulation relates to
 electronic circulation. The relationship is less obvious this time.

```{r linear regressions circulation, echo=FALSE, message=FALSE, warning=FALSE}
kid.lin <- lm(KIDCIRCL ~ TOTCIR, data=action)

summary(kid.lin)

elec.lin <- lm(ELMATCIR ~ TOTCIR, data=action)

summary(elec.lin)

```

Linear regression models. The first one examines the effect total circulation 
numbers have on circulation of children's materials, and the second one looks at
 the relationship total circulation has on the circulation of electronic 
 materials.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The features I examined had fairly straightforward relationships. They 
generally had very strong linear relationships, with the y axis variable rising
 as the x axis variable does so. The most interesting relationships were the 
 ones that didn't follow a strong linear trend! Circulation of electronic 
 materials graphed against total circulation results in a noiser scatterplot 
 than almost any other I looked at. This would be an interesting avenue for 
 investigation, especially by someone interested in the digital divide, or in 
 how people in different states use their library system.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

There is a strong linear regression relationship between the number of visits
and both the hours open and number of programs. Although the log10
transformation didn't reveal anything new, I have a clearer idea of what it
can be useful for. In general, as both number of hours open and number of 
programs offered increases, visit numbers also increase. Conducting a linear
regression shows that both variables are significant, with R-squared values near
1.

As with visit data, the circulation data revealed that the tested variables are
 significant.


### What was the strongest relationship you found?

Most of the relationships I studied were quite strong, but the number of hours
 open influencing the number of visits was particularly powerful.

# Multivariate Plots Section

```{r Visits vs Hours Open by Number of Programs, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(y=VISITS, x=HRS_OPEN), data=physical) +
  geom_point(aes(color=TOTPRO),size= 5) +
  labs(x="No. of Visits", y="No. of Hours Open")


```

This plot shows the relationship between hours open and visits, colored by 
number of programs. 

```{r Linear Regression, echo=FALSE, message=FALSE, warning=FALSE}

lin_all <- lm(VISITS ~ TOTPRO + HRS_OPEN, data=physical)

summary(lin_all)
```

Linear regression analysis for programs and hours open predicting visits.

```{r loanto and loanfm, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=LOANTO, y=LOANFM), data=action) +
  geom_point(aes(color=POPU_ST), size=5) +
  labs(x="Loans Out", y="Loans In")

```

Comparing loan to and loan from, colored by population. The loans numbers are
almost completely identical, creating a linear regression probably close to 1. 
They are not, however, influenced by population size.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#this block transforms the action df to create a tidy one to make barplots.

tidy.action <- gather(action, "type", "n", 'POPU_ST':'LOANFM')
```

```{r Action barplots by state, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 7, fig.width = 9}
ggplot(aes(x=type, y=n), data=tidy.action) +
  geom_bar(stat='identity', aes(fill=type)) +
  facet_wrap(~tidy.action$STABR) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Variable", y="Total Number")
```

A series of barplots, broken down by state, to compare all variables in the 
action table.

```{r Action without total circulation, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 7, fig.width = 9}
action.notot <- subset(action, select = c('STABR', 'ELMATCIR', 'KIDCIRCL', 
                                    'LOANFM', 'LOANTO', 'POPU_ST','REFERENC'))

action.notot <- gather(action.notot, 'type', 'n', 'ELMATCIR':'REFERENC')

ggplot(aes(x=type, y=n), data=action.notot) +
  geom_bar(stat='identity', aes(fill=type)) +
  facet_wrap(~action.notot$STABR) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Variable", y="Total Number")


```

This is the same series of barplots as above, although with total circulation 
removed to better show the relative volumes of the other variables.


```{r Circulation by state, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 7, fig.width = 9}

action.cir <- subset(action, select = c('STABR', 'TOTCIR', 'KIDCIRCL', 'ELMATCIR'))
action.cir <- gather(action.cir, 'type', 'n', 'TOTCIR':'ELMATCIR')

ggplot(aes(x=type, y=n), data=action.cir) +
  geom_bar(stat='identity', aes(fill=type)) +
  facet_wrap(~action.cir$STABR) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Circulation Types", y="Total Number")


```

These plots examine the relative numbers of total circulation (which includes
the other two variables), circulation of kid's books and circulation of 
electronic materials.


```{r pop on circ, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=POPU_ST, y=TOTCIR), data=action) +
  geom_point(aes(color=physical$VISITS), size=5) +
  labs(x="State Population", y="Total Circulation")

```

This scatterplot shows the effects of population size on total circulation 
numbers, colored by number of visits. The outlier with a relatively low 
population and second-highest circulation number is Ohio.

```{r pop on visits, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=POPU_ST,y=VISITS), data=physical) +
  geom_point(aes(color=action$TOTCIR), size=5) +
  labs(x="State Population", y="No. of Visits")

```

Effects of population size on number of visits, colored by total circulation
numbers. As population size increases, so do number of visits.

```{r visits on circ, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=VISITS, y=action$TOTCIR), data=physical) +
  geom_point(aes(color=POPU_ST), size=5) +
  labs(x="No. of Visits", y="Total Circulation")

```

This plot shows the relationship between number of visits and total circulation
 with circulation rising as visits go up. As with the plot above, the 
 population numbers do not map perfectly cleanly.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

I was particularly interested to see that hours open and number of programs 
both influenced number of visits, with hours open having a greater influence. 
Though not exactly surprising (more people will do things at libraries that 
aren't programs, than people who will attend programs), it's nice to have it
broken down.

### Were there any interesting or surprising interactions between features?

Unsurprisingly, circulation is the action most done of the variables examined,
 and circulation of children's materials always outpaces electronic circulation.
  As Americans increase connectivity, I would be very interested to see how 
  that relationship changes, if it ever does.

One thing I was really pleased to learn was that although it was occasionally
 annoying to only have 53 data points for any variable, this makes outliers 
 easy to identify, and set up as case studies or potential deep dives into 
 state-level policies or data that could influence their outlier-ness.

I was really surprised to see the precise, consistent relationship between 
loans to and loans from. It raises the question if there's some kind of policy 
that means the two must equal each other, or does it really naturally work out
 that way? 
 
There's also a really fun outlier when data from both tables is compared, and 
is visible on scatter and barplots. Ohio has unusually high circulation numbers
 for its population and number of visits, and I would love to know what causes 
 that!

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I created several linear regression models. The one that models the effects of 
hours open plus number of programs on visit numbers is particularly interesting 
to me, since it gives a better idea of what is the best predictor.

The biggest limitation is that I really am looking at a very small number of 
variables, especially for something like visit numbers. There are so many 
influences on that, that looking at one or two is more valuable for comparing 
them to each other, than seriously assuming that they have an effect on the 
feature I'm modeling.

For example, visit numbers are likely driven in part by variables not even 
located in these tables -- access to libraries, locations of populations, 
awareness of the library resources, general feelings of welcome, etc.



------

# Final Plots and Summary

### Plot One
```{r Plot_One, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=TOTPRO, y = VISITS), data=physical) +
  geom_point() +
  geom_smooth() +
  xlab('Total Number of Programs') +
  ylab('Total Number of Visits') +
  ggtitle('Relationship between Programs and Visits')

```

### Description One

I chose this plot because it's an excellent representative of most of the 
relationships I examined -- largely linear, with a few interesting outliers. 
With the smoothing, it's clearer in this plot what was discovered in the linear
 modeling: that library programs as a predictor variable does not have a completely
  perfect relationship to visit numbers. That said, when combined with hours 
  open, program numbers is still significant to a p-value of .01, and when 
  assessed alone, its adjusted r-squared value is .8377, a relationship first 
  established by and confirmed by this plot.

### Plot Two
```{r Plot_Two, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 7, fig.width = 9}

ggplot(aes(x=type, y=n), data=action.cir) +
  geom_bar(stat='identity', aes(fill=type), show.legend=F) +
  facet_wrap(~action.cir$STABR) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.text.y = element_text(vjust = .5)) +
  ylab('Circulation Numbers') +
  xlab('Circulation Types') +
  scale_x_discrete(labels = c('Electronic Materials', 'Children\'s Materials', 'Total Materials')) +
  ggtitle('Circulation Numbers by State')


```

### Description Two

I chose this plot as it gives a good overview of the way circulation differs
 from state to state, and lets one pull out the most interesting data at the
  state level. It also shows up the limitations of the dataset -- the radically
   smaller numbers in AS and GU mean that they don't actually offer useful
    information at this scale. It also makes clear that circulation of 
    children's materials always outpaces that of electronic materials, and that 
    often children's materials make up a very significant percentage of overall
     circulation. 

### Plot Three
```{r Plot_Three, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=LOANTO, y=LOANFM), data=action) +
  geom_point(aes(color=POPU_ST, size=10)) +
  xlab('Number of Books loaned to other Libraries') +
  ylab('Number of Books borrowed from other Libraries') +
  labs(color='State Population') +
  stat_smooth(method = 'lm') +
  ggtitle('Book Loans to and From, Colored by Population')

```

```{r include=FALSE}
loanlin <- lm(LOANTO ~ POPU_ST, data=action)
summary(loanlin)
```


### Description Three

I chose this plot because of how eerily regular the linear relationship between
loans to and loans from is. The linear regression line goes through nearly every
 point, suggesting that there is probably an artificial influence on the loan 
 system. I was also interested by the fact that population does not map onto 
 the dataset in the same regular way -- states with very high populations don't
 necessarily see more loans, and the state with the highest number of loans has
  one of the lowest populations. This (lack of) strong relationship was 
  illustrated in the correlation plot at the beginning of the Bivariate Plots
   section.
   
A quick regression calculation, with population predicting loans to other 
libraries resulted in a low adjusted R-squared of .1318, confirming the lack 
of relationship.

------

# Reflection

The most challenging thing was looking at the entire dataset and figuring out
 what questions I wanted to ask. With 124 variables to pick from, there were
  numerous directions I could have gone in. I chose visitor engagement because I
   have the most knowledge about that firsthand, having previously worked in the
    sector. I was able to eventually combine my two dataframes, but to some 
    extent, I wish I had organized them better from the start.
    
I was really pleased at how clear the relationships were between the variables
 I examined -- I feel spoiled by how easy the dataset was, ultimately, to work
 with! It was really interesting to see the linear models, and it gave me a lot
  of ideas of where else this could go, even just with this variable set -- for
   example, the effect of population size. Or, bringing in other variables, the 
   effects of funding, number of librarians, even collection size.
   
I'm really interested in the few outliers in the data -- why does Ohio have
 such a high circulation rate? Why do states with high populations have 
 relatively low loan rates? A lot of the more unusual data was surprising. Also,
  having worked in the sector, I was surprised and pleased to see the
   relationship between number of programs and visits -- when you're desperate
    to get people into your site, it doesn't always feel like programs do very
     much.